---
title: "Grouping and Subqueries"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    css: ../../../styles.css
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center')
```

# Learning Objectives<br>

* Understand `GROUP BY` and `HAVING` for group level aggregation and filtering
* Understand the concept of a subquery and see examples of use

**Duration - 1 hour**<br>

# Setup (again)...

Re-establish the connection to the `omni_pool` database in `DBeaver`.

```{r, eval=TRUE, include=FALSE, message=FALSE, warning=FALSE}
library(RPostgres)

db = dbConnect(
  Postgres(), 
  host = 'db-postgresql-lon1-20135-do-user-7581050-0.b.db.ondigitalocean.com',
  port = 25061,
  dbname = 'omni_pool',
  user = 'omni_user',
  password = 'niytna5mcwsz124i',
  sslmode = 'require',
  bigint = 'numeric'
)
```

# `GROUP BY`

Consider the following problem:

<br>
<center>
**"Find the number of employees within each department of the corporation."**
</center>
<br>

We could solve this manually, but it would be a real pain. First, we would need to get a list of the departments in the corporation (we saw how to do this earlier using `DISTINCT`), and then write a query using a `COUNT()` aggregate with a `WHERE` clause specifying each department in turn, something like:

```{sql, connection = db, output.var="out"}
SELECT 
  COUNT(id) AS number_employees 
FROM employees 
WHERE department = 'Legal';
```
```{r, echo=FALSE, eval=TRUE}
out
```

and so on 11 more times for each department. This is annoying, and not very general `SQL` or maintainable code: we need to know the 12 departments before we start! Thankfully, `SQL` gives us the `GROUP BY` operator to automate the process!

```{sql, connection = db, output.var="out"}
SELECT 
  department, 
  COUNT(id) AS num_employees 
FROM employees 
GROUP BY department;
```
```{r, echo=FALSE, eval=TRUE}
out
```

Yay, this looks more useful! Note what SQL has done here: it first **groups** records by a specified column (`department` in this case), and then applies the aggregate function (`COUNT()`) to **each group**.

Let's see another few examples of grouping in queries:

<br>
<center>
**"How many employees are there in each country?"**
</center>
<br>

```{sql, connection = db, output.var="out"}
SELECT 
  country, 
  COUNT(id) AS num_employees 
FROM employees 
GROUP BY country;
```
```{r, echo=FALSE, eval=TRUE}
out
```

<br>
<center>
**"How many employees in each department work either 0.25 or 0.5 FTE hours?"**
</center>
<br>

```{sql, connection = db, output.var="out"}
--Note the combination of a WHERE clause to filter records and then a GROUP BY. 
--Could also use IN rather than BETWEEN here
SELECT 
  department, 
  COUNT(id) AS num_fte_quarter_half 
FROM employees 
WHERE fte_hours BETWEEN 0.25 AND 0.5 
GROUP BY department;
```
```{r, echo=FALSE, eval=TRUE}
out
```

You'll see here that we are counting the `id` column. You will have seen earlier that we used `COUNT(*)` which counts all rows returned. The difference between the two occurs if there are any `NULL`s in the column you choose for the count.

Let's take a look at the difference using a column which contains some `NULL`s - the `first_name` column:

```{sql, connection = db, output.var="out"}
SELECT 
  COUNT(first_name) AS count_first_name,
  COUNT(id) AS count_id,
  COUNT(*) AS count_star
FROM employees;
```
```{r, echo=FALSE, eval=TRUE}
out
```


The difference in counts arises because there are `NULL` entries in the `first_name` column, and `COUNT()` skips `NULL`s. Generally speaking, if you want a count of rows, stick to using `COUNT(*)` or `COUNT()` of a primary key (which cannot be `NULL` by definition), e.g. `COUNT(id)` in this case.

<br>
<center>
**"Find the longest time served by any one employee in each department."**
</center>
<br>

This is quite a difficult one! We can use the `NOW()` function to return the current date and time, and then calculate the time served by an employee as `NOW()-start_date`. So, the longest time served by any employee is `NOW()-MIN(start_date)`. This is the aggregate function we will apply to the records grouped by department.

```{sql, connection = db, output.var="out"}
SELECT 
  department, 
  NOW()-MIN(start_date) AS longest_time
FROM employees 
GROUP BY department;
```
```{r, echo=FALSE, eval=TRUE}
out
```

This looks broadly like what we're seeking, but notice that the difference `NOW()-MIN(start_date)` gives a confusingly high amount of precision: days, hours, minutes, seconds, etc... Let's get just the days part of this date-time difference. Enter the useful `EXTRACT()` function (you might have used this in the extension questions in yesterday's homework). Let's also see how we could convert this value to years:

```{sql, connection = db, output.var="out"}
SELECT 
  department, 
  EXTRACT(DAY FROM NOW()-MIN(start_date)) AS longest_time_days,
  ROUND(EXTRACT(DAY FROM NOW()-MIN(start_date))/365) AS longest_time_approx_years
FROM employees 
GROUP BY department;
```
```{r, echo=FALSE, eval=TRUE}
out
```

Here's the general rule for `GROUP BY`. Once you've grouped rows, your `SELECT` statement is allowed to retrieve the column specified in `GROUP BY`, together with the result of a aggregate function applied across rows in the group. Let's see what happens when we break this rule by trying to `SELECT` the `first_name` in an earlier query:

```{sql, connection = db, eval=FALSE}
SELECT 
  department, 
  first_name,
  COUNT(id) AS num_employees 
FROM employees 
GROUP BY department
```

You'll get an error: `'ERROR: column "employees.first_name" must appear in the GROUP BY clause or be used in an aggregate function'`!

We like to think of this using an analogy:

* Imagine each row's data has been written on an index card, and we have a big stack of cards for all the rows in the table. 
* A `GROUP BY` operation sorts the cards into separate piles: each pile corresponding to a different value of the `GROUP BY` column (for clarity, we'll assume we `GROUP BY` just one column). 
* We stick a Post-it(TM) on the top of each pile with the label of that group. We're allowed to `SELECT` that label after `GROUP BY`, and we're also allowed to `SELECT` the result of an aggregate function applied to each pile, as this too results in **a single value** that could be easily written on each Post-it. 
* We're **not allowed** to `SELECT` any other column, as this would result in a range of values that couldn't easily be written on each Post-it, because each card in the pile might have a **different value** in that column (remember, all the cards in each pile have in common the value of the `GROUP BY` column).

Don't worry if `GROUP BY` takes a while to understand: it's one of the more difficult concepts in `SQL`. People often struggle a bit with it, but you'll feel more comfortable after some practice. Speaking of which...

<br>
<blockquote class="task">
**Task - 10 mins** 

Write two queries using `WHERE` and `GROUP BY` to solve the following problems:

<br>
<center>
**1. "How many employees in each department are enrolled in the pension scheme?"**
  
**2. "Perform a breakdown by country of the number of employees that do not have a stored first name."**
</center>
<br>

<details>
<summary>**Hints**</summary>  
1. we need to apply a `WHERE` clause on `pension_enrol`.<br>
2. remember the `IS NULL` operator.
</details>

<details>
<summary>**Solution**</summary>
1.  

```{sql, connection = db, output.var="out"}
SELECT 
  department, 
  COUNT(id) as num_employees_pension 
FROM employees 
WHERE pension_enrol = TRUE 
GROUP BY department
```
```{r, echo=FALSE, eval=TRUE}
out
```

2.  

```{sql, connection = db, output.var="out"}
SELECT 
  country, 
  COUNT(id) as num_employees_no_first 
FROM employees 
WHERE first_name IS NULL 
GROUP BY country
```
```{r, echo=FALSE, eval=TRUE}
out
```
</details>
</blockquote>
<br>

# `HAVING`

So far we've seen that the `WHERE` clause let's us filter **rows** (or **records**), but what if we wish to filter **groups** by some value of an aggregate function? This is where the `HAVING` operator comes in!

Imagine one of our earlier queries had been even more specific:

<br>
<center>
**"Show those departments in which at least 40 employees work either 0.25 or 0.5 FTE hours"**
</center>
<br>

```{sql, connection = db, output.var="out"}
SELECT 
  department, 
  COUNT(id) as num_fte_quarter_half 
FROM employees 
WHERE fte_hours IN (0.25, 0.5) 
GROUP BY department 
HAVING COUNT(id) >= 40
```
```{r, echo=FALSE, eval=TRUE}
out
```

We've added a `HAVING` clause **after** the `GROUP BY`: notice that it filters using an **aggregate function** applied to a column of the original table.

Let's try one more:

<br>
<center>
**"Show any countries in which the minimum salary amongst pension enrolled employees is less than 21,000 dollars."**
</center>
<br>

```{sql, connection = db, output.var="out"}
SELECT 
  country, 
  MIN(salary) as min_salary 
FROM employees 
WHERE pension_enrol = TRUE 
GROUP BY country 
HAVING MIN(salary) < 21000
```
```{r, echo=FALSE, eval=TRUE}
out
```

Let's break down the two filtering steps:  

* First, **filter the rows** to include only pension enrolled employees: `WHERE pension_enrol = TRUE`
* Next, group the filtered records by `country`, and then **filter the groups** to include only countries in which the minimum salary is less than $21,000: `HAVING MIN(salary) < 21000`

<br>
<blockquote class="task">
**Task - 5 mins** 
This is about as tough as `SQL` will get for us in this lesson! Write a query using `WHERE`, `GROUP BY` and `HAVING` to solve the following:

<br>
<center>
**"Show any departments in which the earliest start date amongst grade 1 employees is prior to 1991"**
</center>
<br>

<details>
<summary>**Hints**</summary>  

* You can use `MIN(date)` to get the **earliest** value of a date column
* `WHERE` will involve `grade`
* `GROUP BY` will involve department
* `HAVING` will involve `MIN()` of a column

</details>

<details>
<summary>**Solution**</summary>

```{sql, connection = db, output.var="out"}
SELECT 
  department, 
  MIN(start_date) as earliest_start_date_grade_1_pre_91
FROM employees
WHERE grade = 1
GROUP BY department
HAVING MIN(start_date) < '1991-01-01'
```
```{r, echo=FALSE, eval=TRUE}
out
```
</details>
</blockquote>
<br>

# Subqueries

Sometimes we would like to use a value **produced by one query inside another query**. We want to do this **programmatically** (i.e. by code) rather than manually (e.g. by running one query, noting the value, then using that value in a second query). This sounds a bit abstract, but it's actually pretty common! As an example, consider this problem:

<br>
<center>
**"Find all the employees in Japan who earn over the company-wide average salary."**
</center>
<br>

We might think a straightforward way to do this is:

```{sql connection = db, eval = FALSE}
SELECT *
FROM employees
WHERE country = 'Japan' AND salary > AVG(salary)
```

but we get an `ERROR:  aggregate functions are not allowed in WHERE`! 

In fact, we need to use a **subquery** here. This is just **one query embedded inside another**

```{sql, connection = db, output.var="out"}
SELECT *
FROM employees
WHERE country = 'Japan' AND salary > (SELECT AVG(salary) FROM employees)
```
```{r, echo=FALSE, eval=TRUE}
out
```

If it helps, we can think of 'building this query up' in two steps. First, we write what we want to achieve using a place-holder name `average_salary`:  

```{sql, eval=FALSE}
SELECT *
FROM employees
WHERE country = 'Japan' AND salary > average_salary
```

We know we need to use a subquery to get `average_salary`, as it involves an aggregate function, the `AVG(salary)`. So let's write the subquery as a separate query:  

```{sql, eval=FALSE}
SELECT AVG(salary)
FROM employees"
```

Now wherever we see `average_salary` in the first query, we replace it with the subquery in parentheses

```{sql, connection = db, output.var="out"}
SELECT *
FROM employees
WHERE country = 'Japan' AND salary > (SELECT AVG(salary) FROM employees)
```
```{r, echo=FALSE, eval=TRUE}
out
```

<br>
<blockquote class="task">
**Task - 5 mins**

Write a query using a subquery to answer the following:

<br>
<center>
**"Find all the employees in Legal who earn less than the mean `salary` in that same department."**
</center>
<br>

<details>
<summary>**Solution**</summary>
```{sql, connection = db, output.var="out"}
SELECT *
FROM employees
WHERE department = 'Legal' AND salary < (
  SELECT AVG(salary)
  FROM employees
  WHERE department = 'Legal'
)
```
```{r, echo=FALSE, eval=TRUE}
out
```
Notice the use of indentation to mark out the subquery as being separate from the outer query: we recommend following this in your own code.
</details>
</blockquote>
<br>

<br>
<blockquote class="task">
**Harder Task - 10 mins** 

This is quite hard! Combine what you've learned in this lesson to answer the following:

<br>
<center>
**"Find all the employees in the United States who work the most common full-time equivalent hours across the corporation."**
</center>
<br>

Think of this as being two separate tasks combined:

* Write a query to find the most common full-time equivalent hours across the corporation
* Use this as a subquery in an outer query to find those employees in the United States who work those hours

<details>
<summary>**Hints**</summary>
The first query corresponds to the **mode** in statistics: the most commonly occurring value in a set of values. Think of grouping the employees by `fte_hours`, counting the number of employees in each group, and then finding the `fte_hours` with the largest count. Can you extend your subquery to work for ties?
</details>

<details>
<summary>**Solution**</summary>

*Query for the mode*

```{sql, connection = db, output.var="out"}
SELECT fte_hours
FROM employees
GROUP BY fte_hours
ORDER BY COUNT(*) DESC
LIMIT 1
```
```{r, echo=FALSE, eval=TRUE}
out
```

*Query for the mode - accounting for ties*

In `PostgreSQL` when you use a subquery in the `FROM` clause, you **must** give it an alias, hence `AS temp` below.

```{sql, connection = db, output.var="out"}
SELECT fte_hours
FROM employees
GROUP BY fte_hours
HAVING COUNT(*) = (
  SELECT MAX(count)
  FROM (
    SELECT COUNT(*) AS count
    FROM employees
    GROUP BY fte_hours
  ) AS temp
)
```
```{r, echo=FALSE, eval=TRUE}
out
```

*Use the mode query as a subquery in outer query*

```{sql, connection = db, output.var="out"}
SELECT *
FROM employees
WHERE country = 'United States' AND fte_hours IN (
  SELECT fte_hours
  FROM employees
  GROUP BY fte_hours
  HAVING COUNT(*) = (
    SELECT MAX(count)
    FROM (
      SELECT COUNT(*) AS count
      FROM employees
      GROUP BY fte_hours
    ) AS temp
  )
)
```
```{r, echo=FALSE, eval=TRUE}
out
```

</details>
</blockquote>
<br>

# Recaps

## Order of definition

| Order of definition | Keyword | Specifies | Required? |
| --- | --- |--- | --- |
| 1 | `SELECT` | Column to query | Yes |
| 2 | `AS` | Column alias | No |
| 3 | `FROM` | Table to query |  Yes |
| 4 | `WHERE` | Row-level filter | No |
| 5 | `GROUP BY` | Grouping for aggregates | No |
| 6 | `HAVING` | Group-level filter | No |
| 7 | `ORDER BY` | Sort order | No |
| 8 | `LIMIT` | How many records to return | No |

## Useful mnemonic for order of definition

| Keyword | Mnemonic |
| --- | --- |
| `SELECT AS` | So |
| `FROM` | Few |
| `WHERE` | Workers
| `GROUP BY` | Go
| `HAVING` | Home
| `ORDER BY` | On Time

## Order of execution

The order above applies for **writing** `SQL` queries, but it is also helpful to understand the **order of execution** of the parts of a `SQL` query. This can help explain why, for example, an alias that you declare in the `SELECT` part can't be used in `HAVING`, and so on...

| Order of execution | Keyword |
| --- | --- |
| 1 | `FROM` |
| 2 | `WHERE` |
| 3 | `GROUP BY` |
| 4 | `HAVING` |
| 5 | `SELECT` |
| 6 | `ORDER BY` |
| 7 | `LIMIT` |



