---
title: "Manipulating Returned Data"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    css: ../../../styles.css
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center')
```

# Learning Objectives<br>

* Be able to create column aliases using `AS`
* Use `DISTINCT` to return unique records by column
* Understand and be able to use aggregate functions
* Be able to sort records and limit the number returned

**Duration - 1 hour 45 minutes**<br>

This is a continuation of the last in which we learned how to filter data at the row level using the `WHERE` clause. Now we'll add to our query toolkit with some more advanced syntax! 

# Setup (again)...

Re-establish the connection to the `omni_pool` database in `DBeaver` (you may not have to do this if you ran a query recently).

```{r, eval=TRUE, include=FALSE, message=FALSE, warning=FALSE}
library(RPostgres)

db = dbConnect(
  Postgres(), 
  host = 'db-postgresql-lon1-20135-do-user-7581050-0.b.db.ondigitalocean.com',
  port = 25061,
  dbname = 'omni_pool',
  user = 'omni_user',
  password = 'niytna5mcwsz124i',
  sslmode = 'require',
  bigint = 'numeric'
)
```

# Manipulating returned data

After we have applied conditions to filter data in a `SELECT` statement, we can also manipulate what is returned. The easiest way to do this is by limiting the returned columns (fields)!

```{sql, connection = db, output.var="out"}
SELECT 
  id, 
  first_name, 
  last_name 
FROM employees 
WHERE department = 'Accounting';
```
```{r, echo=FALSE, eval=TRUE}
out
```


So here we return only the `id`, `first_name` and `last_name` columns.

`SQL` offers additional operators to manipulate the returned data, which we will examine in this lesson.

# Aliases via `AS`

<br>
<center>
**"Can we get a list of all employees with their first and last names combined together into one field called 'full_name'?"**
</center>
<br>

**Column aliases** are the way to solve problems like these! First, we use the `CONCAT()` function to **concatenate** (this is just a fancy way of saying 'join strings together') each pair of names into the full name. Next, we set up a column alias using `AS full_name` to store the concatenated strings. 

Let's see this in operation first with some strings where the contents are known:

```{sql, connection = db, output.var="out"}
SELECT
  CONCAT('Hello', ' ', 'there!') AS greeting;
```
```{r, echo=FALSE, eval=TRUE}
out
```

You can also use the `||` operator for this:

```{sql, connection = db, output.var="out"}
SELECT
  'Hello' || ' ' || 'there!' AS greeting;
```
```{r, echo=FALSE, eval=TRUE}
out
```

We concatenated `'Hello'` with a space `' '` and `'there!'`, and named the column `greeting`. You can always try things like this in `SQL` to explore and help you understand syntax and functions. Now we try a similar approach with the columns in our table:

```{sql, connection = db, output.var="out"}
SELECT 
  id, 
  first_name, 
  last_name, 
  CONCAT(first_name, ' ', last_name) AS full_name 
FROM employees;
```
```{r, echo=FALSE, eval=TRUE}
out
```

The new `full_name` column will be at the right of the output. We see a problem with this: some of the records now have single names for `full_name`, but this represents a problem with the underlying data. We could add in a `WHERE` clause to filter out these problem rows.


<br>
<blockquote class="task">
**Task - 2 mins** 

Add a `WHERE` clause to the query above to filter out any rows that don't have both a first and second name.  

<details>
<summary>**Hint**</summary>
You need an `AND` combination of two `IS NOT NULL` conditions
</details>

<details>
<summary>**Solution**</summary>
```{sql, connection = db, output.var="out"}
SELECT 
  id, 
  first_name,
  last_name,
  CONCAT(first_name, ' ', last_name) AS full_name 
FROM employees 
WHERE first_name IS NOT NULL AND last_name IS NOT NULL;
```
```{r, echo=FALSE, eval=TRUE}
out
```
</details>
</blockquote>
<br>

It is good practice to use aliases when creating new columns or using aggregate functions (coming up soon), so that if someone else (including your future self) examines your `SQL` output, the columns will have useful names. To see what happens without an alias, let's run the query again without `AS full_name`:

```{sql, connection = db, output.var="out"}
SELECT 
  id, 
  first_name,
  last_name,
  CONCAT(first_name, ' ', last_name)
FROM employees 
WHERE first_name IS NOT NULL AND last_name IS NOT NULL;
```
```{r, echo=FALSE, eval=TRUE}
out
```

# `DISTINCT()`

<br>
<center>
**"Our database may be out of date! After the recent restructuring, we should now have six departments in the corporation. How many departments do employees belong to at present in the database?"**
</center>
<br>

How do we solve this problem? It's not enough to simply return **all** the departments of the employees, as there will be a large amount of duplication. Instead, we need the **unique** list of departments. The `DISTINCT()` function returns a unique list.

```{sql, connection = db, output.var="out"}
SELECT 
  DISTINCT(department) 
FROM employees;
```
```{r, echo=FALSE, eval=TRUE}
out
```


# Aggregate functions

<br>
<center>
**"How many employees started work for the corporation in 2001?"**
</center>
<br>

The `COUNT()` aggregate function can help us with counting problems:

```{sql, connection = db, output.var="out"}
SELECT 
  COUNT(*) 
FROM employees 
WHERE start_date BETWEEN '2001-01-01' AND '2001-12-31';
```
```{r, echo=FALSE, eval=TRUE}
out
```

In addition to `COUNT()`, we have the following aggregate functions:
<br>

| Function | Purpose |
| --- | --- |
| `SUM()` | sum of a column |
| `AVG()` | average of a column |
| `MIN()` | minimum value of a column |
| `MAX()` | maximum value of a column |

As mentioned above it's good practice to use an alias (using the `AS` function) to give meaning to the result, such as:

```{sql, connection = db, output.var="out"}
SELECT 
  COUNT(*) AS total_employ_2001
FROM employees 
WHERE start_date BETWEEN '2001-01-01' AND '2001-12-31';
```
```{r, echo=FALSE, eval=TRUE}
out
```


<br>
<blockquote class="task">
**Task - 10 mins** 

Design queries using aggregate functions and what you have learned so far to answer the following questions:<br>

<br>
<center>
**1. "What are the maximum and minimum salaries of all employees?"**

**2. "What is the average salary of employees in the Human Resources department?"**

**3. "How much does the corporation spend on the salaries of employees hired in 2018?"**
</center>
<br>

<details>
<summary>**Solution**</summary>
1.  

```{sql, connection = db, output.var="out"}
SELECT 
  MAX(salary) AS max_salary
FROM employees;
```
```{r, echo=FALSE, eval=TRUE}
out
```

```{sql, connection = db, output.var="out"}
SELECT 
  MIN(salary) AS min_salary
FROM employees;
```
```{r, echo=FALSE, eval=TRUE}
out
```

We can also do this in a single query:  

```{sql, connection = db, output.var="out"}
SELECT 
  MAX(salary) AS max_salary,
  MIN(salary) AS min_salary
FROM employees;
```
```{r, echo=FALSE, eval=TRUE}
out
```

2.  

```{sql, connection = db, output.var="out"}
SELECT 
  AVG(salary)as avg_sal
FROM employees 
WHERE department = 'Human Resources';
```
```{r, echo=FALSE, eval=TRUE}
out
```

3.
```{sql, connection = db, output.var="out"}
SELECT 
  SUM(salary) AS total_sals_2018
FROM employees 
WHERE start_date BETWEEN '2018-01-01' AND '2018-12-31';
```
```{r, echo=FALSE, eval=TRUE}
out
```
</details>
</blockquote>
<br>

# Sorting by columns

The `ORDER BY` operator lets us **sort** the returns of queries, either in descending (`DESC`) or ascending (`ASC`) order. The `ORDER BY` operator and associated keywords **always comes after** any `WHERE` clause.

The `LIMIT` operator is a natural partner to `ORDER BY`: it lets us limit **how many** records are returned by a query.

So, we saw before the minimum and maximum salaries of employees in the corporation. Let's find out which employees earn those salaries using the new operators!

```{sql, connection = db, output.var="out"}
--First, the minimum salaried employee. 
--Order employees by salary in *ascending* order, return the first, 
--We need to be careful with NULLs here, let's filter out employees with NULL salary
SELECT * 
FROM employees 
WHERE salary IS NOT NULL 
ORDER BY salary ASC 
LIMIT 1;
```
```{r, echo=FALSE, eval=TRUE}
out
```

```{sql, connection = db, output.var="out"}
--Next, the maximum salaried employee. 
--Order employees by salary in *descending* order, return the first 
SELECT * 
FROM employees 
WHERE salary IS NOT NULL 
ORDER BY salary DESC 
LIMIT 1;
```
```{r, echo=FALSE, eval=TRUE}
out
```

You see above we had problems in ordering with `NULL`s in the column we're sorting on. We can either filter them out with a `WHERE` clause, as we did above, or we can use the operators `NULLS FIRST` or `NULLS LAST` to specify where to put them in the list of records. These operators are always placed immediately after `DESC` or `ASC` for the respective column.

Let's rewrite the queries above using these operators:

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
ORDER BY salary ASC NULLS LAST 
LIMIT 1;
```
```{r, echo=FALSE, eval=TRUE}
out
```

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
ORDER BY salary DESC NULLS LAST 
LIMIT 1;
```
```{r, echo=FALSE, eval=TRUE}
out
```

Finally, we can perform **multi-level sorts** on two or more columns:

<br>
<center>
**"Order employees by full-time equivalent hours, highest first, and then alphabetically by last name."**
</center>
<br>

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
ORDER BY 
  fte_hours DESC NULLS LAST, 
  last_name ASC NULLS LAST;
```
```{r, echo=FALSE, eval=TRUE}
out
```

<br>
<blockquote class="task">
**Task - 5 mins** 

Write queries to answer the following questions using the operators introduced in this section.

<br>
<center>
**1. "Get the details of the longest-serving employee of the corporation."**
  
**2. "Get the details of the highest paid employee of the corporation in Libya."**
</center>
<br>

<details>
<summary>**Solution**</summary>

1.  

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
ORDER BY start_date ASC NULLS LAST 
LIMIT 1;
```
```{r, echo=FALSE, eval=TRUE}
out
```

2.  

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
WHERE country = 'Libya' 
ORDER BY salary DESC NULLS LAST 
LIMIT 1;
```
```{r, echo=FALSE, eval=TRUE}
out
```

</details>
</blockquote>
<br>

**A note on ties**

Be careful when you're trying to find rows that have the **maximum** or **minimum** in a particular column in situations where **ties are likely** (i.e. more than one row has the maximum or minimum value). In this case using `LIMIT 1` will just arbitrarily select **one of the tied rows**, while you would usually rather see **all** tied rows. 

A better approach in this case would involve two queries:

* Write a first query to find the maximum or minimum value in a column as required
* Use this maximum or minimum value in the `WHERE` clause of a second query to find **all** rows with that value

We'll cover the concept of **subqueries** tomorrow, which will let us combine these 'two queries' into a single query.

# Recaps

## Order of definition

We have *nearly* completed our discussion of query syntax, here are **all** the different components of a `SELECT` query, the order in which they must appear, and whether they are required or optional <br>

| Order of definition | Keyword | Specifies | Required? |
| --- | --- |--- | --- |
| 1 | `SELECT` | Column to query | Yes |
| 2 | `AS` | Column alias | No |
| 3 | `FROM` | Table to query |  Yes |
| 4 | `WHERE` | Row-level filter | No |
| 5 | `GROUP BY` | Grouping for aggregates | No |
| 6 | `HAVING` | Group-level filter | No |
| 7 | `ORDER BY` | Sort order | No |
| 8 | `LIMIT` | How many records to return | No |

We will discuss the `GROUP BY` and `HAVING` keywords tomorrow!

## Useful mnemonic for order of definition

| Keyword | Mnemonic |
| --- | --- |
| `SELECT AS` | So |
| `FROM` | Few |
| `WHERE` | Workers
| `GROUP BY` | Go
| `HAVING` | Home
| `ORDER BY` | On Time

## Order of execution

The order above applies for **writing** `SQL` queries, but it is helpful to understand the **order of execution** of the parts of a `SQL` query. This can help explain why, for example, an alias that you declare in the `SELECT` part can't be used in `HAVING`, and so on...

| Order of execution | Keyword |
| --- | --- |
| 1 | `FROM` |
| 2 | `WHERE` |
| 3 | `GROUP BY` |
| 4 | `HAVING` |
| 5 | `SELECT` |
| 6 | `ORDER BY` |
| 7 | `LIMIT` |



