---
title: "R Notebook"
output: html_notebook
---

# Unit testing in R
## why test in R - reproducibility, consistancy

# Leap year assignment
```{r}
is_leap_year <- function(year){
  #add in condition for non numeric data
  if (!is.numeric(year)) {
    stop("Non-numeric input :(")
  }
    
  if (year %% 100 == 0){
      if (year %% 400 == 0){
        return(TRUE)
      } else {
        return(FALSE)
      }
  }


if (year %% 4 == 0){
  return(TRUE){
    else{
      return(FALSE)
    }
  }
}
```

```{r}
is_leap_year <- function(year){
  
  #add in condition for non-numeric data
  if (!is.numeric(year)) {
    stop("Non-numeric input :(")
  }
  
  if (year %% 100 == 0){
    if (year %% 400 == 0){
      return(TRUE)
    } else {
      return(FALSE)
    }
  }
  
  if (year %% 4 == 0){
    return(TRUE)
  } else {
    return(FALSE)
  }
  
}
```


```{r}
is_leap_year <- function(year){
  
  if (year %% 100 == 0){
    if (year %% 400 == 0){
      return(TRUE)
    } else {
      return(FALSE)
    }
  }
  
  if (year %% 4 == 0){
    return(TRUE)
  } else {
    return(FALSE)
  }
  
}


```

## it's dividing by 100 and checking for a 

```{r}
library(testthat)

```

expect_equal(): checking that the objects give the answer we expect
expect_error(): checking thtat code throws an error

```{r}
a <- 10
```

```{r}
expect_equal(a, 10)
```

```{r}
expect_equal(a, 11)
```
test_that()

```{r}
test_that("Real leap years return TRUE", {
  expect_equal(is_leap_year(2020), TRUE)
  expect_equal(is_leap_year(2000), TRUE)
})
```

```{r}
# second test : one expectation
test_that("Character input returns an error", {
  expect_error(is_leap_year("year"))
})
## Test passed 🎉
```

## the test na,e is "Character input returns an error"
##it's looking to see if the text is contained in the code, when it sees that year isn't there it says - yes that is an error (ie it's not there) and says yes, it was a successful error

```{r}
test_that("Character input returns an error", {
  expect_error(is_leap_year(2010))
})
```

```{r}
test_that("Non-numeric input returns an error", {
  expect_error(is_leap_year("year"))
  expect_error(is_leap_year(TRUE))
})
```

```{r}
test_that("check that 2017 and 1900 are not leap years",{
  expect_equal(is_leap_year(2017), FALSE)
  expect_equal(is_leap_year(1900), FALSE)
})
```

# assertive programming
```{r}
library(tidyverse)
weather_one <- read.csv("data/delhi_weather_one.csv")

weather_one
```

```{r}
weather_report <- function(weather){
  # Our average temperature, humidity and wind speed 
  weather_means <- 
  weather %>% 
  summarise(
    mean_temp = mean(temp),
    mean_humidity = mean(humidity),
    mean_windspeed = mean(wind_speed)
  )

# Show hot and humid days
hot_and_humid <- 
weather %>% 
  filter(temp > 20) %>% 
  filter(humidity > 55)

return(list(weather_means, hot_and_humid))
}

weather_report(delhi_weather_one)
```

```{r}
weather_two <- read_csv("data/delhi_weather_two.csv")

weather_report(weather_two)
```

```{r}
install.packages("assertr")
```

# using assertr

```{r}
library(assertr)
```

```{r}
weather_report <- function(weather){
  
  # put in the checks on the data first
  weather %>% 
    # humidiyt is a proportion
    verify(humidity >= 0 & humidity <= 100)
    #lowest ever temp in Delhi was -31, and hightes +51
    verify(temp >= -35 & temp <= 55) %>% 
    verify(wind_speed >= 0 & wind_speed <= 108)
  
  # Our average temperature, humidity and wind speed 
  weather_means <- 
  weather %>% 
  summarise(
    mean_temp = mean(temp),
    mean_humidity = mean(humidity),
    mean_windspeed = mean(wind_speed)
  )

# Show hot and humid days
hot_and_humid <- 
weather %>% 
  filter(temp > 20) %>% 
  filter(humidity > 55)

return(list(weather_means, hot_and_humid))
}

weather_report(weather_two)
view(weather_two)
```

## example with stopifnot for more general chunks

```{r}
weather_report <- function(weather){

  stopifnot(
    nrow(weather) > 15,
    ncol(weather) == 4,
    drop_na(weather) == weather
  )  
  # put in the checks on the data first
  weather %>% 
    # humidiyt is a proportion
    verify(humidity >= 0 & humidity <= 100)
    #lowest ever temp in Delhi was -31, and hightes +51
    verify(temp >= -35 & temp <= 55) %>% 
    verify(wind_speed >= 0 & wind_speed <= 108)
  
  # Our average temperature, humidity and wind speed 
  weather_means <- 
  weather %>% 
  summarise(
    mean_temp = mean(temp),
    mean_humidity = mean(humidity),
    mean_windspeed = mean(wind_speed)
  )

# Show hot and humid days
hot_and_humid <- 
weather %>% 
  filter(temp > 20) %>% 
  filter(humidity > 55)

return(list(weather_means, hot_and_humid))
}

weather_report(weather_two)

```

```{r}

```

