---
title: "R Notebook"
output: html_notebook
---

# Coding Best Practice

R Script
start at top with all the librarys eg

library(RXKCD)
library(beepr)

cmd + shft + r = section break

# Task on rewriting badly written code
original code:

AmountSpent=c(9.1,8.55,14.32,14.17,14.34,7.97,9.79,11.47,12.44)
Category=c('Lunch'
              , 'Dinner'
              , 'Lunch'
              , 'Lunch'
              , 'Cinema'
              , 'Drinks'
              , 'Lunch'
              , 'Lunch'
              , 'Dinner')
my.week.of.spending=tibble(AmountSpent=AmountSpent, SpendingCategory=Category)
str(my.week.of.spending)
View(my.week.of.spending)
my.week.of.spending
summarise(my.week.of.spending,sum(AmountSpent))
summarise(filter(my.week.of.spending,SpendingCategory=='Lunch'),sum(AmountSpent))

amended:

AmountSpent = c(9.1, 8.55, 14.32, 14.17, 14.34, 7.97, 9.79, 11.47, 12.44)
Category = c('Lunch', 'Dinner', 'Lunch', 'Lunch', 'Cinema', 'Drinks', 'Lunch', 'Lunch', 'Dinner')
my_week_of_spending = tibble(AmountSpent = AmountSpent, SpendingCategory = Category)
str(my_week_of_spending)
View(my_week_of_spending)
my_week_of_spending
summarise(my_week_of_spending, sum(AmountSpent))
summarise(filter(my_week_of_spending, SpendingCategory == 'Lunch'),sum(AmountSpent))

```{r}
# run styler

AmountSpent <- c(9.1, 8.55, 14.32, 14.17, 14.34, 7.97, 9.79, 11.47, 12.44)
Category <- c("Lunch", "Dinner", "Lunch", "Lunch", "Cinema", "Drinks", "Lunch", "Lunch", "Dinner")
my_week_of_spending <- tibble(AmountSpent = AmountSpent, SpendingCategory = Category)
str(my_week_of_spending)
View(my_week_of_spending)
my_week_of_spending
summarise(my_week_of_spending, sum(AmountSpent))
summarise(filter(my_week_of_spending, SpendingCategory == "Lunch"), sum(AmountSpent))
```

```{r}
library(dplyr)
library(MASS)

# this will identify any conflict between packages
conflicts()
library(tidyverse)
library(dplyr)

```

```{r}
mtcars %>% 
  dplyr::select(mpg, hp) %>% 
  filter(mpg < 50) %>% 
  summarise(mean = mean(hp))
```

```{r}
install.packages("errorst")
library("errorist")


```

```{r}
some_letters <- c("E", "R", "R", "O", "R")
```

